Creating Something Different: Similarity, Contrast, and Representativeness in Categorization

### 2024-02-22 Manuscripts/cbb2023/sn-article.tex
work to get paper compiling as it does in overleaf
getting towards compiling in normal latex environment
Minor progress on revision

\jyear{2023}%

@@

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate} 
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

@@

\abstract{The ability to generate new concepts and ideas is among the most fascinating aspects of human cognition, but we do not have a strong understanding of the
cognitive processes and representations underlying concept generation. In this
paper, we study the generation of new categories using the computational and
behavioral toolkit of traditional artificial category learning. Previous work in
this domain has focused on how the statistical structure of known categories
generalizes to generated categories, overlooking whether (and if so, how)
contrast between the known and generated categories is a factor. We report three
experiments demonstrating that contrast between what is known and what is
created is of fundamental importance for categorization. We propose two novel
approaches to modeling category contrast: one focused on exemplar dissimilarity
and another on the representativeness heuristic. Our experiments and
computational analyses demonstrate that both models capture different aspects of
contrast's role in categorization.}

@@

The second hypothesis for how contrast might affect categorization is contrast
as {\em representativeness}. The representativeness heuristic
\citep{kahneman1972subjective} states that people make judgments regarding an
outcome based on how representative it is of the evidence given in the current
context. The heuristic is a powerful theoretical construct that has been used to
capture a range of complex patterns of human judgments
\citep{kahneman73,tversky74,tversky83}, especially those that deviate from
normative theory (as given by a straightforward application of Bayes' rule). For
example, the coinflip sequence TTHTH is perceived to be more random (due to it
being representative of a coinflip sequence) than TTTTT (despite both having the
same probability of having been generated from a fair coin). Although there is a
healthy debate as to whether perceived randomness {\em really} deviates from
normative theory \citep{griffiths18,hahn09}, the representativeness heuristic
remains one of the main explanations of human judgments \citep{reimers18}. To
formalize the hypothesis that people expect a new category to be representative
of the opposite of the current category, or contrast as {\em
  representativeness}, we use the Bayesian formulation of representativeness
\citep{tenenbaum01}, which has been used successfully to capture human
performance in color categorization \citep{abbott16}, image category learning
\citep{abbott11}, and language grammar learning \citep{rafferty10}. We do so by
extending the state-of-the-art category generation model
\citep{jern2013probabilistic} and find that human category generation can
also be captured as generating representative, rather than probable, samples.

The outline of the article is as follows. First we describe previous
computational formalizations of theories of category generation and empirical
work investigating them. We then present two hypotheses for how contrast might
affect categorization and formalize them in computational models. The first is a
novel exemplar model, which is designed to generate categories that
systematically differ from existing categories in the domain. The second assumes
that the goal of category generation is to create representative samples of the
opposite of the observed categories. We present two experiments demonstrating
strong and systematic effects of category contrast on concept generation, and we
qualitatively and quantitatively analyze the behavioral results and the performance of each model in
capturing human category generation. A third experiment is performed to
highlight the differences between the two contrast models. We conclude with a
discussion of the implications of our results for categorization and directions
for future work.

@@

\begin{figure}
	\begin{center} 
    \includegraphics[width=0.95\textwidth]{fig1.eps}
      \caption{The effect of weights on PACKER generation of a category `B' example, following exposure
        to one member of category `A' and one member of category `B'.
        Predictions are shown for three different parameterizations (differing
        in values of $\theta_t$ and $\theta_c$): {\em (a)}
        Predictions based on contrast similarity only. {\em (b)}
        Predictions based on target similarity only. {\em (c)} Predictions with
        both constraints considered.}
		\label{fig:packer-examples}
	\end{center}
\end{figure}

@@

\begin{figure}
	\begin{center} \includegraphics{fig2.eps}
		\caption{Generation of a category `B' example, following exposure
to one member of category `A' and one member of category `B'. Predictions are
shown for (a) the original hierarchical Bayesian model and (b) the
representativeness model.}
		\label{fig:representative-examples}
	\end{center}
\end{figure}

@@

\begin{figure}
    \begin{center} \includegraphics[width=0.8\textwidth]{fig13.eps}
    \caption{PACKER generation of a category `B' example, following exposure to
one member of category `A' and one member of category `B'. Predictions are shown
for different attention settings: {\em (a)} Increased weighting of the X-axis.
{\em (b)} Increased weighting of the Y-axis. {\em (c)} Uniform weighting
(identical to Figure \ref{fig:packer-examples}).}
    \label{fig:packer-attention}
    \end{center}
\end{figure}

>>>>

\jyear{2024}%

@@

@@

\abstract{The ability to generate new concepts and ideas is among the most fascinating aspects of human cognition,
but we do not have a strong understanding of the
cognitive processes and representations underlying concept generation. Previous work in
this domain has focused on how the statistical structure of known categories
generalizes to generated categories, overlooking whether (and if so, how)
contrast between the known and generated categories is a factor. In this paper, 
we explore a different factor: contrast from known categories. We propose two novel
approaches to modeling category contrast: one focused on exemplar dissimilarity
and another based on the representativeness heuristic. Across three behavioral experiments, 
we find that people generate new categories that contrast from observed categories and 
distribute exemplars acoss "unoccupied" regions of stimulus space. The model based on 
the representativeness heuristic captured human category generation better when 
the known category was well captured by a Gaussian distribution. Conversely, the 
exemplar-based model captured human generated categories better when the known category
was not Gaussian distributed. Our results suggest contrast is a fundamental principle used in
generating exemplars of a new category.}

@@

The second hypothesis for how contrast might affect categorization is contrast
as {\em representativeness}. The representativeness heuristic
\citep{kahneman1972subjective} states that people make judgments regarding an
outcome based on how representative it is of the evidence given in the current
context. The heuristic is a powerful theoretical construct that has been used to
capture a range of complex patterns of human judgments
\citep{kahneman73,tversky74,tversky83}, especially those that deviate from
normative theory (as given by a straightforward application of Bayes' rule). For
example, the coinflip sequence TTHTH is perceived to be more random (due to it
being representative of a coinflip sequence) than TTTTT (despite both having the
same probability of having been generated from a fair coin). Although there is a
healthy debate as to whether perceived randomness {\em really} deviates from
normative theory \citep{griffiths18,hahn09}, the representativeness heuristic
remains one of the main explanations of human judgments \citep{reimers18}. To
formalize the hypothesis that people expect a new category to be representative
of the opposite of the current category, or contrast as {\em
  representativeness}, we use the Bayesian formulation of representativeness
\citep{tenenbaum01}, which has been used successfully to capture human
performance in color categorization \citep{abbott16}, image category learning
\citep{abbott11}, and language grammar learning \citep{rafferty10}. We do so by
extending a Bayesian category generation model
\citep{jern2013probabilistic} and find that human category generation can
also be captured as generating representative, rather than probable, samples.

The outline of the article is as follows. First we describe previous
computational formalizations of theories of category generation and empirical
work investigating them. We then present two hypotheses for how contrast might
affect categorization and formalize them in computational models. The first is a
novel exemplar model, which is designed to generate categories that
systematically differ from existing categories in the domain. The second assumes
that the goal of category generation is to create representative samples of the
opposite of the observed categories. We present two experiments demonstrating
strong and systematic effects of category contrast on concept generation, and we
analyze the behavioral results and the performance of each model in
capturing human category generation at the individual-level. Based on the model-
based analyses, we develop and conduct a third experiment that distinguishes the two
computational formulations of contrast. The third experiment found the exemplar-based
version of contrast to capture human performance better. We conclude with a
discussion of the implications of our results for categorization and directions
for future work.

@@

\begin{figure}
	\begin{center} 
    \includegraphics[width=0.95\textwidth]{figs/fig1.eps}
      \caption{The effect of weights on PACKER generation of a category `B' example, following exposure
        to one member of category `A' and one member of category `B'.
        Predictions are shown for three different parameterizations (differing
        in values of $\theta_t$ and $\theta_c$): {\em (a)}
        Predictions based on contrast similarity only. {\em (b)}
        Predictions based on target similarity only. {\em (c)} Predictions with
        both constraints considered.}
		\label{fig:packer-examples}
	\end{center}
\end{figure}

@@

\begin{figure}
	\begin{center} \includegraphics{figs/fig2.eps}
		\caption{Generation of a category `B' example, following exposure
to one member of category `A' and one member of category `B'. Predictions are
shown for (a) the original hierarchical Bayesian model and (b) the
representativeness model.}
		\label{fig:representative-examples}
	\end{center}
\end{figure}

@@

\begin{figure}
    \begin{center} \includegraphics[width=0.8\textwidth]{figs/fig13.eps}
    \caption{PACKER generation of a category `B' example, following exposure to
one member of category `A' and one member of category `B'. Predictions are shown
for different attention settings: {\em (a)} Increased weighting of the X-axis.
{\em (b)} Increased weighting of the Y-axis. {\em (c)} Uniform weighting
(identical to Figure \ref{fig:packer-examples}).}
    \label{fig:packer-attention}
    \end{center}
\end{figure}
